---
title: "Data Wrangling"
author: "Rowana Ahmed"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stringr)
library(xlsx)
```

## Read in Data

```{r,  warning = FALSE}
facilities = read_csv('data/mammography_facilities.csv')
screen_rate = read_csv('data/500_Cities__City.csv')
chdb_metrics = read_csv('data/CHDB_data_city_allv10_1.csv')
```


## High Level: National Facilities Mapping

objective: merge following data sources -  

* mammography facilities

* screen rates (mammography, papsmear, core preventive services)

* breast cancer deaths

Step 1: subset 500 cities data to relevant screening metrics -> **screen_data** df
```{r}
names(screen_rate)
#test
# # Relevant measures (each have corresponding age adjusted measures): 
# # "Papanicolaou smear use among adult women aged 21–65 Years" - PAPTEST
# # "Mammography use among women aged 50–74 Years" - MAMMOUSE 
# # "Older adult women aged >=65 Years who are up to date on a core set of clinical preventive services: Flu shot past Year, PPV shot ever, Colorectal cancer screening, and Mammogram past 2 Years" - COREW
# 
screen_metrics <- c('StateAbbr', 'PlaceName', 'Population2010',
                   'COREW_CrudePrev', 'COREW_Crude95CI', 
                   'COREW_AdjPrev', 'COREW_Adj95CI',
                   'MAMMOUSE_CrudePrev', 'MAMMOUSE_Crude95CI', 
                   'MAMMOUSE_AdjPrev', 'MAMMOUSE_Adj95CI',
                   'PAPTEST_CrudePrev', 'PAPTEST_Crude95CI', 
                   'PAPTEST_AdjPrev', 'PAPTEST_Adj95CI')

screen_data <-  screen_rate %>%
                select(screen_metrics) %>% rename(state_abbr = StateAbbr,
                                                  city_name = PlaceName)

```

Step 2: subset city health dashboard metrics to target variables -> **factors_data** df
```{r}
names(chdb_metrics)
unique(chdb_metrics$metric_name)

# Relevant measures

# ## environmental
# "Air pollution - particulate matter" - 2
# "Housing with potential lead risk" - 12
# 
# ## socioeconomic
# "High school completion" - 10
# "Income Inequality" - 13
# "Limited access to healthy foods" -15
# "Teen births"  - 33 
# "Walkability" - 27
# "Uninsured"  - 25
# "Unemployment - annual, neighborhood-level" - 24
# "Racial/ethnic diversity"  - 21
# 
# ## individual health
# "Binge drinking"  - 3
# "Diabetes" - 6
# "Frequent mental distress" -7
# "Frequent physical distress" -8
# "Prenatal care" - 34
# "Low birthweight" - 35
# "Breast cancer deaths" - 28
# "Smoking" - 22
# "Obesity" - 17

indiv_health <- list(3, 6, 7, 8, 34, 35, 28, 22, 17)
socioecon <- list(10, 13, 15, 33, 27, 25, 24, 21)
env <- list(2,12)

factors_data <- chdb_metrics %>% filter(metric_number %in% append(append(indiv_health, 
                                                                  socioecon), env)) %>%
                 select(c(state_abbr, city_name, metric_name, metric_number, data_yr_type,
                          group_name, group_number, 
                          num, denom, est, lci, uci))

factors_data <- factors_data %>% mutate(metric_name = paste(metric_name, group_name, sep='_'))



```


Step 3: aggregate facilities count per city -> **site_counts** df
```{r}
facilities <- facilities %>% mutate(state = str_to_upper(state)) %>% 
                             mutate(city = str_to_title(city))
site_counts <- facilities %>% group_by(state, city) %>% count(city)
names(site_counts) <- c('state_abbr', 'city_name', 'number_sites')

head(site_counts)


```

Step 4: Merge higher level data into single df


```{r}

wide_factors <- factors_data %>% pivot_wider(id_cols = c(state_abbr, city_name),
                                             names_from = metric_name,
                                             values_from = est)

temp <- left_join(screen_data, wide_factors, by=c('state_abbr', 'city_name'))
city_total_data <- left_join(temp, site_counts, by=c('state_abbr', 'city_name'))

write.xlsx(city_total_data, 'data/city_total_data.xlsx')

```


## Lower Level Geography Focused Data