---
title: "facilities_latlong_nyc"
author: "Dougie Zubizarreta"
date: "11/27/2020"
output: html_document
---

# Load packages
```{r}
library(tidyverse)
library(mapproj)
library(sf)
```

# Read in the mammography facilities .csv file containing all facilities in the US
```{r}
mammography_facilities <- read_csv("data/mammography_facilities.csv")
```

# Filter to create a new dataframe with only facilities in New York state
```{r}
table(mammography_facilities$state)

mammography_facilities_ny <- mammography_facilities %>%
  filter(state == "NY")
mammography_facilities_ny

table(mammography_facilities_ny$state)
```

# Filter to create a new dataframe with only facilities in New York city
```{r}
table(mammography_facilities_ny$city)

mammography_facilities_nyc <- mammography_facilities_ny %>%
  filter(city %in% c("Staten Isalnd","Staten Island","brooklyn","Brooklyn", "BROOKLYN", "Bronx","Jamaica","Jackson Heights","Richmond Hill","Rego Park","Ozone Park","NEW YORK","New York","Long Island City","Laurelton","Kew Gardens","Howard Beach","Glendale","Forest Hills","Flushing","Far Rockaway","Elmhurst","Cambria Heights","Bayside","Astoria"))
mammography_facilities_nyc

table(mammography_facilities_nyc$city)
```

# Rename some of the values to make sure the spelling is correct
```{r}
mammography_facilities_nyc = mammography_facilities_nyc %>%
  mutate(city = recode(city, 
                          `brooklyn`="Brooklyn",
                          `BROOKLYN` = "Brooklyn",
                          `NEW YORK` = "New York",
                          `Staten Isalnd` = "Staten Island"
                          ))
table(mammography_facilities_nyc$city)
```

# Group the values by borough (i.e., Manhattan, Brooklyn, Queens, the Bronx, Staten Island)
```{r}
colnames(mammography_facilities_nyc)[colnames(mammography_facilities_nyc) == "city"] <- "borough"

mammography_facilities_nyc = mammography_facilities_nyc %>%
  mutate(borough = recode(borough, 
                          `Astoria`="Queens",
                          `Bayside` = "Queens",
                          `Cambria Heights` = "Queens",
                          `Elmhurst` = "Queens",
                          `Far Rockaway`="Queens",
                          `Flushing` = "Queens",
                          `Forest Hills` = "Queens",
                          `Glendale` = "Queens",
                          `Howard Beach`="Queens",
                          `Jackson Heights` = "Queens",
                          `Jamaica` = "Queens",
                          `Kew Gardens` = "Queens",
                          `Laurelton`="Queens",
                          `Long Island City` = "Queens",
                          `Ozone Park` = "Queens",
                          `Rego Park` = "Queens",
                          `Richmond Hill` = "Queens",
                          ))
table(mammography_facilities_nyc$borough)
```

# Export to .csv file. Manually enter in latitude and longitude coordinates using Google Maps. Add these latitude and longitude coordinates into new columns in the .csv file. 
```{r, eval = FALSE}
write.csv(mammography_facilities_nyc,"data/mammography_facilities_nyc.csv", row.names = FALSE)
```

# Import the updated .csv file containing the latitude and longitude coordinates.
```{r}
mammography_facilities_nyc_latlong <- read_csv("mammography_facilities_nyc_latlong.csv")
```

# Read in the 2010 NYC census tracts shape file from NYC OpenData 
[Visit NYC OpenData!](https://data.cityofnewyork.us/City-Government/2010-Census-Tracts/fxpq-c8ku).
```{r}
census_tracts <- st_read("data/geo_export_afcfe2e2-6376-4d66-b290-605e0ac77ee5.shp", quiet = TRUE)
```

# Create a plot of facilities with mammography services in New York City
```{r}
p <- census_tracts %>%
  select(geometry) %>%
  plot(main = "Facilities Offering Mammography Services in NYC", lwd = 0.25)

coord <- mapproject(mammography_facilities_nyc_latlong$longitude, mammography_facilities_nyc_latlong$latitude, projection = "", parameters = NULL, orientation = NULL)

p + points(coord, pch = 20, cex = 0.5, col = "red")
```
